---
title: "Historical Markers"
description: "Contribution to week 27 TidyTuesday: visualizing historical markers"
---

# Historical Markers

```{r}

library('tidyverse')
library('camcorder')
library('tidycensus')
library('here')

here::i_am("historical-markers.qmd")

```

First, we load the data. The marker data set comes from the *Historical Marker Database USA Index*, provided via the #TidyTuesday forum. I also collected county population data from the *{tidycensus}* package. Lastly, to achieve the tile map, I loaded a shape set generated from a package I'm designing that makes custom tiled maps.

```{r}

# Load marker data

utah_markers <- 
  readr::read_csv(
    'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-07-04/historical_markers.csv'
  ) %>%
  filter(state_or_prov == "Utah")

# Load county population data

county_pop <-
  tidycensus::get_acs(
    geography = "county", 
    variables = "B01003_001", 
    state = "UT", 
    geometry = FALSE,
    year = 2020
  )

# Load county tile data

load(here("utah_county_tiles.rds"))

```

I next manipulated some of the data:

*   Filtering for Utah state markers
*   Summarizing the count of markers by county
*   Cleaning up the county names to facilitate joining the data sets
*   Joining the disparate data
*   Creating the "per 1,000 residents" variable

```{r}

grouped_markers <- 
  utah_markers %>%
  select(title, county_or_parish) %>%
  mutate(county_or_parish = stringr::str_replace(county_or_parish, " County", "")) %>%
  group_by(county_or_parish) %>%
  summarise(count = n()) %>%
  ungroup()

county_adj <- 
  county_pop %>%
    mutate(county = stringr::str_replace(NAME, " County, Utah", "")) %>%
    select(county, estimate)

utah_counties <- 
  inner_join(
    uc_tiles, 
    grouped_markers, 
    by = c("id" = "county_or_parish")
  ) %>%
  inner_join(
    county_adj, 
    by = c("id" = "county")
  ) %>%
  mutate(
    per_1k = (count / estimate) * 1000
  )

```

Next, I setup the camcorder to create a high res, nicely scaled visualization.

```{r}

gg_record(
  dir = 'C:/Users/Adam Bushman/Pictures/Camcorder', 
  device = 'png', 
  width = 16, 
  height = 16, 
  units = "cm", 
  dpi = 300
)

```

I also setup the colors and scales. Using *{RColorBrewer}* and some other custom functions, I created a gradient variable for the background and a variable for the contrasting text. 

```{r}

my_colors = list(
  high = "#ff930f", 
  low = "#183182", 
  background = "#FFFFFF"
)

color_scale = colorRampPalette(c(my_colors$low, my_colors$high))
gradient_colors = color_scale(nrow(utah_counties))

map_gradient <- function(val, min, max) {
  scaled_value <- (val - min) / (max - min)
  
  index <- floor(scaled_value * (length(gradient_colors) - 1)) + 1
  
  gradient_colors[index]
}

utah_counties <-
  utah_counties %>%
  mutate(
    background_color = map_gradient(per_1k, min(per_1k), max(per_1k)), 
    text_color = purrr::map_chr(background_color, adamb::contrast_text_color)
  )

```

Finally, I setup the the title, subtitle, and caption texts with the appropriate markdown styling and then constructed the plot.

I added several annotations to highlight the extremes of population adjusted historical marker concentration.

```{r}

my_title = glue::glue(
  "Utah state population trends **have not**<br>", 
  "aligned with <span style='color:{my_colors$high}'>**historical markers**</span>"
)

my_subtitle = glue::glue(
  "Number of markers in each county per 1,000 residents as of 2020 US census<br><span style='color:{my_colors$high}'>**HIGH**</span> and <span style='color:{my_colors$low}'>**LOW**</span> concentration of population adjusted markers"
)

my_caption = paste0(
  "Source: Historical Marker Database USA Index & {tidycensus} package\n", 
  "Designed by Adam Bushman (@adam_bushman)"
)

ggplot() +
  geom_polygon(
    aes(x, y, group = id, fill = background_color), 
    data = utah_counties, 
    show.legend = FALSE
  ) +
  geom_text(
    aes(center_x, center_y, label = id, color = text_color), 
    data = utah_counties %>% select(-c(x, y)), 
    size = 2.75
  ) +
  scale_fill_identity() +
  scale_color_identity() + 
  scale_x_continuous(expand = c(0,0)) +
  labs(
    title = my_title, 
    subtitle = my_subtitle,
    caption = my_caption
  ) +
  # Salt Lake County
  annotate(
    "segment", x = 27, y = 54.5, xend = 85, yend = 54.5, 
    lwd = 0.5, color = "black"
  ) +
  annotate(
    "point", x = 27, y = 54.5, size = 1.5, color = "black"
  ) +
  annotate(
    "text", x = 85, y = 54.5, hjust = 1, vjust = -2, 
    size = 4, color = "black", lineheight = 0.8, fontface = "bold", 
    label = "Salt Lake County"
  ) +
  annotate(
    "text", x = 85, y = 54.5, hjust = 1, vjust = -0.75, 
    size = 3.5, color = "black", lineheight = 0.8, 
    label = "Population outpaced markers"
  ) +
  annotate(
    "text", x = 74, y = 54.5, hjust = -0.1, vjust = 1.45, 
    size = 4, color = "black", lineheight = 2, 
    label = "markers\nresidents"
  ) +
  annotate(
    "text", x = 74, y = 54.5, hjust = 1, vjust = 1.1, 
    size = 10, color = my_colors$low, lineheight = 0.8, 
    fontface = "bold", label = "158\n~1M"
  ) +
  # Wayne County
  annotate(
    "segment", x = 27, y = 21.5, xend = 85, yend = 21.5, 
    lwd = 0.5, color = "black"
  ) +
  annotate(
    "point", x = 27, y = 21.5, size = 1.5, color = "black"
  ) +
  annotate(
    "text", x = 85, y = 21.5, hjust = 1, vjust = -2, 
    size = 4, color = "black", lineheight = 0.8, fontface = "bold", 
    label = "Wayne County"
  ) +
  annotate(
    "text", x = 85, y = 21.5, hjust = 1, vjust = -0.75, 
    size = 3.5, color = "black", lineheight = 0.8, 
    label = "Markers outpaced population"
  ) +
    annotate(
    "text", x = 74, y = 21.5, hjust = -0.1, vjust = 1.45, 
    size = 4, color = "black", lineheight = 2, 
    label = "markers\nresidents"
  ) +
  annotate(
    "text", x = 74, y = 21.5, hjust = 1, vjust = 1.1, 
    size = 10, color = my_colors$high, lineheight = 0.8, 
    fontface = "bold", label = "10\n~3K"
  ) +
  theme_void() +
  theme(
    plot.title = ggtext::element_markdown(size = 22, lineheight = 1), 
    plot.subtitle = ggtext::element_markdown(size = 12, lineheight = 1), 
    plot.caption = element_text(size = 8, hjust = 0), 
    plot.background = element_rect(fill = my_colors$background, color = NA), 
    
    plot.margin = margin(0.5, 1, 0.5, 1, "cm")
  )

```



